Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents

### 2025-07-03 doc/cicm25-paper/rebuttal.txt

We thank the reviewers for their thorough and thoughtful comments and suggestions.

We concur with the reviewers that the paper is somewhere between a research paper and a
system description, and we have struggled with this in the paper writing process. Indeed
we had started out as the former, and then realized that there is more to the topic than a
system description could convey. We did not try to bloat the text to make it more suitable
to the category, but tried to layout the task of semantic authoring -- which we see as a
novel category -- as a general problem to be addressed by the CICM community.

### 2025-07-23 doc/cicm25-paper/paper.tex
reworked paper

\documentclass[runningheads]{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{paralist}
\usepackage[hidelinks]{hyperref}
\usepackage{orcidlink}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\usepackage[noabbrev]{cleveref}
\usepackage{tikz}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{docicon}

@@

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\newcommand\defianno{\textsf{defianno}\xspace}
\newcommand\FTML{\textsf{FTML}\xspace}
\newcommand\WOIDE{\textsf{WOIDE}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}

\title{
    % Semantic Authoring in the \sTeX/\ALeA World
    % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents
}
\titlerunning{
    Semantic authoring in a flexiformal context \textellipsis
}

@@

  In this paper we discuss the nature of the semantic authoring problem and argue that,
  while sharing aspects of classical (informal) authoring and formal authoring, it is a
  problem of its own that needs special editing support facilities.  We present and
  evaluate a simple utility that supports the process of annotating technical terms with
  semantic references to a flexiformal content commons that enhances annotation
  productivity by almost an order of magnitude.
\end{abstract}
% \setcounter{tocdepth}{3}
% \tableofcontents

\section{Introduction}
Arguably, dealing with large document collections is one of the key factors in the knowledge-driven society and economy.
There are currently two main contenders for machine support in this area.
Symbolic/logic-based technologies and sub-symbolic/machine-learning-based AI, e.g.\ via LLMs or chatbots.
They have complementary strengths and challenges:
ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
-- in particular because of the cost of semantic authoring, which we discuss in this paper.

@@

\paragraph{Overview \& Contribution}
In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
Using \sTeX/\ALeA as an exemplary case study,
we will provide an overview of the authoring support (\Cref{sec:tools})
and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\Cref{sec:snify}).
Afterwards, we discuss on-going work to integrate \snify better
into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).

@@

Before we discuss semantic authoring in general, let us introduce a concrete example from
the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the challenges
involved and to ground the solution we propose.

@@

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
\end{figure}

Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both). Concretely this is about converting the {\LaTeX} string

\begin{lstlisting}[numbers=left,firstnumber=3,
caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated item:
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
In the latter, the author has annotated the word ``terminal'' with the concept it refers
to with via the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.

In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}: words of the domain jargon used to refer to the concept or
object.  Verbalizations can vary between between subdomains and communities of
practice. In our example, the verbalization -- i.e. the natural language phrase used to
express the symbol in English -- ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in
\Cref{lst:losa}).

The annotation refers to a definition in the domain model module shown in
\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
\lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in
line 3.

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

@@

To annotate the word ``terminal'' in the document, the author has to 
\begin{inparaenum}[\em i\rm)]
\item be aware that it is a
technical term, 
\item that the module from \Cref{fig:state-space} exists,
\item know the symbol URI, 
\item manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
  directive if it is not already (recursively) implied and possibly removing directives
  that become redundant by the new one.
\end{inparaenum}
At ca 150,000 words in a semester's worth of lecture notes and ca. 10--15\% of technical
terms a rather daunting task.

@@

For documents in a language $L$,
where $L$ is not English, the annotation practices are a bit
more complex still, in particular because the annotation catalog is usually incomplete.

Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
\sTeX already supports multilinguality:
\begin{itemize}
\item Documents have explicit language annotations (usually in form of the file extension
  \lstinline[mathescape]|.$\llangle lang\rrangle$.tex| where $\llangle lang\rrangle$ is the
  ISO-639 language identifier).
\item Modules in the domain model allow $L$-translations that share the formal
  declarations with another module. In particular all definienda in $L$-modules are
  $L$-specific.
\end{itemize}

But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
practice, as
\begin{enumerate}[\em i\rm)]
\item $L$-verbalizations and symbol names do not often coincide and
\item the domain model may not even have $L$-modules -- especially for the ``smaller''
  languages.
\end{enumerate}
 In this situation, we usually have to invest in
establishing a suitable catalog e.g. by providing $L$-translations (or at least dictionary
information) in the domain model or delineating modules and annotating definienda that the
domain model does not cover in the documents to be annotated.

@@

\paragraph{Formal Authoring}
In programming, the content commons is given formally in the form of the libraries of the
respective programming language, correspondingly the authored content -- a program -- is
also formal and intended for processing by a computer, for which the formality is a
prerequisite. Formal authoring is usually supported by an integrated development
environment (IDE) that makes use of the fully formal content and offers services like
semantic highlighting, tab completion, jumping to variable declarations, function
definitions, and interface definitions, as well as refactoring support. All of these rely
on a syntactic/semantic analysis of the formal content. The situation is very similar for
authoring of formal knowledge, e.g. in program verification or the formalization of
mathematics.

@@

A distinctive feature of semantic authoring that is not apparent at the first glance is
that the added investment in annotation moves the sweet spot in the re-use/customize
trade-off towards sharing and interoperability. In our experience sufficiently annotated
documents can be automatically customized if the semantics has multiple
notations/verbalizations/presentations. Consider for instance course materials on complex
arithmetic, which become re-usable across the electrical engineering/mathematics border if
the imaginary unit in formulae is represented e.g. by a semantic macro
\lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths) and
$j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
reusable/shareable document fragments organized in central or federated repositories and
hubs. This is quite similar to the formal authoring arena, where we see a tendency towards
massively shared math libraries in formal systems and package/library hubs for programming
libraries. As a consequence, semantic authoring support systems should facilitate the
logistics of using and curating shared libraries.

@@

On the formal side, there is a custom VSCode IDE plugin for \sTeX \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.  But such features
only support existing annotations and are, therefore, of limited use when creating new
annotations, which arguably is the most time consuming aspects of semantic authoring.
% \ednote{also: SPARQL API}

@@

The \snify\footnote{ From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash
    sn}'' is a frequently used \sTeX short-hand macro for term annotations.}  utility
\cite{stextools:git} is a simple command-line tool that creates a catalog of
symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the catalog.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation workflow.  The interface
is inspired by traditional spell checkers like \lstinline|ispell|~\cite{ispellman}.  \Cref{fig:snify} shows
a typical situation.

@@

\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the
top-level lecture notes; then the session consists of all (transitively) included
files. Other productivity features include a focus mode that can be used to only annotate
a particular word in the rest of the file, session, or the whole \sTeX/\ALeA corpus before
resuming regular annotation. This reduces the cognitive overhead from switching between
different words and symbol lists.
% Of course the include management still needs to be done locally.

\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to consider different \LaTeX\ macros to identify definienda (e.g.\
\lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
through them as \snify does.  For each potential definiendum, the user can then use fuzzy
search to find the corresponding symbol in the domain model.  This populates the catalog
with verbalizations relevant to the document, making subsequent \snify runs more
effective, especially for documents in languages that are not well covered by the domain
model (i.e.\ not English).

\paragraph{Evaluation}
% \section{Practical Evaluation and Future Work}\label{sec:eval}
\begin{wrapfigure}{r}{0.4\textwidth}
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
\end{wrapfigure}
In our experience, \snify's step-through workflow for annotating term references for symbols
from the domain model is almost an order of magnitude more efficient than writing the
annotations and imports by hand, even for annotators who are familiar with the domain
model.
Given the sheer number of term annotations (see \Cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

Finally, the current interface is not well-suited for on-the-fly annotation while
authoring.
For that, the underlying information (the symbol/verbalization pairs harvested
from the domain model) can be integrated into any IDE. In fact we plan to do this for the
next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.
For now, the typical workflow of early adopters is to first write the document (e.g.\ a set of quiz questions) without annotations, and then run \snify to annotate the new content.

\section{Scaling \snify to the Flexiformal Content Commons }\label{sec:scaling}
One practical problem with \snify system as presented so far is that it only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
ecosystem grows.

The general situation is that on the one hand there is a constantly growing, shared
\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
files -- that are regularly converted to \FTML (\underline{F}lexiformal \underline{T}ext
\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
other hand, authors are working on
\begin{compactenum}[\em i\rm)]
\item a set of local working copies of the archives -- with the intention of publishing
  them on MathHub.info -- and their dependencies (they need to be local since
  \textsf{pdflatex} cannot deal with remote files) and
\item possibly a set of private archives (e.g.\ papers under development or exam
  problems/solutions).
\end{compactenum}
This situation is familiar from software development and also from e.g. the \LaTeX{}
ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
from both.

To cope with this situation, we are working on extending the brand-new \flams system
\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
from the flexiformal content commons would be just a simple SPARQL query and some
postprocessing of the result (de-duplication, stemming, and pairing) away. Note that
\flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
and exported from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}. There are two
more systems that could profit from this: The CPoint system that allows annotating MS
PowerPoint presentations with the \sTeX ontology and can already generate a precursor of
\FTML \cite{Kohlhase:SemanticInteractionDesignDiss:biblatex}, and a semantic extension of
Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
FTML instead of plain HTML.

>>>>

\documentclass[runningheads]{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{paralist}
\usepackage[hidelinks]{hyperref}
\usepackage{orcidlink}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\usepackage[noabbrev]{cleveref}
\usepackage{tikz}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{docicon}
\usepackage{soul}
\def\edited#1{\hl{#1}}

@@

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\newcommand\defianno{\textsf{defianno}\xspace}
\newcommand\FTML{\textsf{FTML}\xspace}
\newcommand\WOIDE{\textsf{WOIDE}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}
\def\flams{\textsf{FLAMS}\xspace}

\lstset{literate=--1}

\title{
    % Semantic Authoring in the \sTeX/\ALeA World
    % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
        Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents
}
\titlerunning{
    Semantic Authoring in a Flexiformal Context \textellipsis
}

@@

  In this paper we discuss the nature of the semantic authoring problem and argue that,
  while sharing aspects of classical (informal) authoring and formal authoring, it is a
  problem of its own that needs special editing support facilities.  We present and
  evaluate a simple utility that supports the process of annotating technical terms with
    semantic references to a flexiformal content commons \edited{and increases} annotation
  productivity by almost an order of magnitude.
\end{abstract}
% \setcounter{tocdepth}{3}
% \tableofcontents

\section{Introduction}
Arguably, handling large document collections is a key factor in our knowledge-driven society and economy.
There are two main contenders for machine support in this area:
symbolic/logic-based technologies and \edited{statistical}/machine-learning-based AI, e.g.\ via LLMs or chatbots.
They have complementary strengths and challenges:
ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
Symbolic technologies, on the other hand, \edited{offer precision and explainability%
%out of the box%
, but face scalability challenges}
-- in particular because of the cost of semantic authoring, which we discuss in this paper.

@@

\paragraph{Overview and Contribution}
In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
Using \sTeX/\ALeA as an exemplary case study,
we will provide an overview of the authoring support (\Cref{sec:tools})
and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\Cref{sec:snify}).
Afterwards, we discuss on-going work to integrate \snify better
into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).

@@

Before we discuss semantic authoring in general, let us introduce a concrete example from the \sTeX/\ALeA world.
    % We will use it to sharpen our intuitions of the challenges involved and to ground the solution we propose.

@@

\begin{figure}[h]\centering
  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
\end{figure}

\begin{figure}[h]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both). Concretely this is about converting the unannotated {\LaTeX} string

\begin{lstlisting}[numbers=left,firstnumber=3,
% caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los
]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated item
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
% caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa
]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
\edited{
where the author has annotated the word ``terminal'' with the concept it refers
to via} the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
In \sTeX, \edited{concepts} are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a \textbf{symbol name}.
 % Every symbol can have one or more verbalizations. %: words of the domain jargon used to refer to the concept or
\edited{
A \textbf{verbalization} is the natural language rendering of a concept.
% object.
Verbalizations are, of course, language-dependent and can vary between subdomains and communities of
practice.
}
In our example, the verbalization ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
% in \Cref{lst:losa}).

The annotation refers to a definition in the domain model module shown in
\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
\lstinline|search-problem?goal state| introduced by the \lstinline|\symdecl*| declaration in
line 3.

@@

To annotate the word ``terminal'' in the document, the author has to 
\begin{inparaenum}[\em i\rm)]
\item be aware that it is a
technical term, 
\item that the module from \Cref{fig:state-space} exists,
\item know the symbol URI, 
\item manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
  directive if it is not already (recursively) implied and possibly removing directives
  that become redundant by the new one.
\end{inparaenum}
\edited{A rather daunting task for a lecture with 50,000 -- 100,000 words in notes and slides and ca. 10--15\,\% of them technical terms.\footnote{
    \edited{
        Estimates based on some of our own lectures materials
        ranging from a small introductory lecture (ca. 30,000 words)
        to a large two-semester lecture on AI (ca. 140,000 words).
        The 10--15\,\% estimate appears to be relatively consistent across our lectures
        but depends on the author's intuition about what a technical term is.
        Other lecturers or lectures on other topics may have very different values.
    }
}}

@@

\edited{The annotation catalog is language-specific.}
% For documents in a language $L$,
% where $L$ is not English, the annotation practices are a bit
% more complex still, in particular because the annotation catalog is usually incomplete \edited{for $L$}.
% Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
\sTeX already supports multilinguality:
\begin{itemize}
\item Documents have explicit language annotations (usually in form of the file extension
  \lstinline[mathescape]|.$\llangle lang\rrangle$.tex| where $\llangle lang\rrangle$ is the
  ISO-639 language identifier).
\item \edited{Modules in the domain model allow translations that share the formal
    declarations with another module.}
        % In particular all definienda in $L$-modules are $L$-specific.
\end{itemize}
% But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
\edited{In practice, the annotation catalog for languages other than English is often inadequate as}
\begin{enumerate}[\em i\rm)]
    \item \edited{the domain model may not have translations in that language -- especially for the ``smaller'' languages, and}
    \item \edited{symbol names, usually English words, cannot be used as a verbalizations.}
\end{enumerate}
\edited{In this situation, we usually have to invest in
establishing a suitable catalog -- e.g. by providing translations (or at least dictionary
information) in the domain model or
delineating modules and annotating definienda that the
domain model does not cover in the documents to be annotated.
}

@@

\paragraph{Formal Authoring}
In programming, the content commons is given formally in the form of the libraries of the
respective programming language. Correspondingly the authored content -- a program -- is
also formal and intended for processing by a computer, for which the formality is a
prerequisite. Formal authoring is usually supported by an integrated development
environment (IDE) that makes use of the fully formal content and offers services like
semantic highlighting, tab completion, jumping to variable declarations, function
definitions, and interface definitions, as well as refactoring support. All of these rely
on a syntactic/semantic analysis of the formal content. The situation is very similar for
authoring of formal knowledge, e.g. in program verification or the formalization of
mathematics.

@@

A distinctive feature of semantic authoring that is not apparent at the first glance is
that the added investment in annotation moves the sweet spot in the re-use/customize
trade-off towards sharing and interoperability. In our experience sufficiently annotated
documents can be automatically customized if the semantics has multiple
notations/verbalizations/presentations. Consider for instance course materials on complex
arithmetic, which become re-usable across the electrical engineering/mathematics border if
the imaginary unit in formulae is represented e.g. by a semantic macro
\lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths) and
$j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
reusable/shareable document fragments organized in central or federated repositories and
hubs. This is quite similar to the formal authoring arena, where we see a tendency towards
massively shared math libraries in formal systems and package/library hubs for programming
libraries. Semantic authoring support systems should therefore facilitate the
logistics of using and curating shared libraries.

@@

On the formal side, there is a custom VSCode IDE plugin for \sTeX \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.  But such features
only support existing annotations and are, therefore, of limited use when creating new
annotations, which arguably is the most time consuming aspect of semantic authoring.
% \ednote{also: SPARQL API}

@@

The \snify\footnote{ From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash
    sn}'' is a frequently used \sTeX short-hand macro for term annotations.}  utility
\cite{stextools:git} is a simple command-line tool that creates a catalog of
symbol-verbalization pairs (see also \Cref{sec:workflows}), analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the catalog.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation workflow.  The interface
is inspired by traditional spell checkers like \lstinline|ispell|~\cite{ispellman}.  \Cref{fig:snify} shows
a typical situation.

@@

\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the
top-level lecture notes; then the session consists of all (transitively) included
files. Another productivity feature is a focus mode that can be used to only annotate
a particular word in the rest of the file, session, or the whole \sTeX/\ALeA corpus before
resuming regular annotation. This reduces the cognitive overhead from switching between
different words and symbol lists.
% Of course the include management still needs to be done locally.

\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents.}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to consider different \LaTeX\ macros to identify definienda (e.g.\
\lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
through them as \snify does.  For each potential definiendum, the user can then use fuzzy
search to find the corresponding symbol in the domain model.  This populates the catalog
with verbalizations relevant to the document, making subsequent \snify runs more
effective, especially for documents in languages that are not well covered by the domain
model (i.e.\ not English).

\paragraph{Evaluation}
% \section{Practical Evaluation and Future Work}\label{sec:eval}
\begin{wrapfigure}{r}{0.4\textwidth}
    \vskip-1.2cm
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
    \vskip-0.2cm
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
  \vskip-0.7cm
\end{wrapfigure}
In our experience, \snify's step-through workflow for annotating term references 
% for symbols from the domain model 
is almost an order of magnitude more efficient than writing the
annotations and imports by hand.\footnote{
    \edited{
        The precise productivity gain depends on the annotator's
        familiarity with the domain model (for the right symbol
        is time-consuming) but also on other factors like
        the typing speed.
        Measurements are complicated by the fact that
        authors tend to annotate more terms when using \snify.
        We have observed speed-ups over IDE-based annotation
        ranging from 3 (for an expert knowing the domain model by heart)
        to 13 (for a beginner).
        The focus mode further increases productivity.
    }
}
% , even for annotators who are familiar with the domain model.
Given the sheer number of term annotations (see \Cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

Finally, the current interface is not well-suited for on-the-fly annotation while
authoring.
For that, the underlying information (the symbol/verbalization pairs harvested
from the domain model) can be integrated into any IDE. In fact we plan to do this for the
next version of the \sTeX plugin for VSCode \cite{sTeX-IDE:git}.
For now, the typical workflow of early adopters is to first write the document (e.g.\ a set of quiz questions) without annotations, and then run \snify to annotate the new content.

\section{Scaling \snify to the Flexiformal Content Commons }\label{sec:scaling}
One practical problem with the \snify system as presented so far is that it only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
ecosystem grows.

The general situation is that on the one hand there is a constantly growing, shared
\textbf{content commons} of public math archives -- currently 200 with about 14\;000 \sTeX
files -- that are regularly converted to \FTML (\underline{F}lexiformal \underline{T}ext
\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
other hand, authors are working on
\begin{compactenum}[\em i\rm)]
\item a set of local working copies of the archives -- with the intention of publishing
  them on MathHub.info -- and their dependencies (they need to be local since
  \textsf{pdflatex} cannot deal with remote files) and
\item possibly a set of private archives (e.g.\ papers under development or exam
  problems/solutions).
\end{compactenum}
This situation is familiar from software development and also from e.g. the \LaTeX{}
ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
from both.

% To cope with this situation, we are working on extending the brand-new \flams system
% \cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
%   here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
% so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
% from the flexiformal content commons would be just a simple SPARQL query and some
% postprocessing of the result (de-duplication, stemming, and pairing) away.
\edited{To cope with this situation, we are planning to extend the brand-new} \flams \edited{system, which among other things powers} \url{https://MathHub.info},
\edited{so that it provides the data necessary to create a
symbol/verbalization catalog.}
Note that
\flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
and exported from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}. There are two
more systems that could profit from this: The CPoint system that allows annotating MS
PowerPoint presentations with the \sTeX ontology and can already generate a precursor of
\FTML \cite{Kohlhase:SemanticInteractionDesignDiss:biblatex}, and a semantic extension of
Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
\FTML instead of plain HTML.
